---
title: "Are we more civilized today?"
author: "Emilie  Andersen"
date: " created 4 October 2021, updated 14 March 2025"
output:
  html_document:
    toc: true         
    toc_depth: 2  
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE)
```
#  Explore Homicide rates in NW Europe
In this exercise, you will load a Manuel Eisner's dataset to capture the developments in homicides from Middle Ages to 20th century. 

**Your task: Explore the data and create the visualisations as instructed in the script.** 

- **Submit to Github this script with solutions and its knitted version.**  
- **Submit to Brightspace a document with the (1) URL of this script and its knitted version in Github (2) answer of 250 words to the Final Question and (3) the Final Plot.**



This exercise is based on the dataset provided by [OurWorldInData project](https://ourworldindata.org/homicides) based at the Oxford University.

## Long-Term Trends in Homicides

Understanding how homicide rates have changed prior to the modern era requires the help of historians and archivists. Manuel Eisner, a criminology professor at the University of Cambridge, and his colleagues published the Historical Violence Database \href{https://cjrc.osu.edu/research/interdisciplinary/hvd}: a compilation of data on long-term trends in homicide rates, in addition to qualitative information such as the cause of death, perpetrator and victim. This database is limited to countries with relatively complete historical records on violence and crime – mainly Western Europe and the US.

Starting in the second half of the nineteenth century, these European regions have consistent police records of those accused of murder or manslaughter and annual counts of homicide victims. To go back further in time, reaching as far back as the thirteenth century, Eisner collected estimates (from historical records of coroner reports, court trials, and the police) of homicide rates made in over ninety publications by scholars.

Homicide rates – measured as the number of homicides per 100,000 individuals – up to 1990 are sourced from Eisner’s (2003) publication and the Historical Violence Database. 

Are homicide rates in Europe today lower or higher than in the past? Using the provided dataset, display and describe the long-run homicide rates for the five European regions: Italy, England, Germany, Netherlands and Scandinavia.

```{r library}
library(tidyverse)
```

## Loading Data

You should always interrogate the source of your data. Who compiled it, from where, what is missing, how representative the data are? Check the data/Metadata.txt to learn about the data provenance.

```{r load-data}
Western_Europe <- read_csv("data/homicide-rates-across-western-europe.csv")
```

## Data Inspection

How clean and analysis-ready is the dataset? Do you understand what the column names represent? What is the difference between rate and homicide number?

```{r inspect}
head(Western_Europe)

```

Ok, the data look good except for the column `Homicide rate in Europe over long-term (per 100,000)` which is not very easy to work with.

- Use the `names()` function and assignment key to relabel this column to `homicides_per_100k` 

```{r relabel column}
names(Western_Europe)[4] <- "homicides_per_100k"

glimpse(Western_Europe)
```

Now, that you have looked at what the data looks like and what it represents, and streamlined it, let's see what big picture it contains.

## Homicide Trends Over Time - Creating Line Plots

- use `ggplot()` function and remember the`+` at the end of the line
- chose a meaningful `geom_......()`  for geometry (hint: points are not great)
- load `Year` on the `x` axis and `homicides_per_100k` column in y axis
- to color individual country entries consistently, assign the country column to the argument `color`.
- provide meaningful title and axis labels
- remember to change the `eval` flag so that the code chunk renders when knitted 

```{r plot-homicides, eval = FALSE}
ggplot(Western_Europe, aes(x = Year, y = homicides_per_100k, color = Entity)) +
  geom_line() +
  labs(
    title = "Long-term Homicide Rates in Western Europe",
    x = "Year",
    y = "Homicides per 100,000 People",
    color = "Country"
  ) +
  theme_bw()

ggsave("data/LongTermHomicideRatesInWesternEurope.png")
```
Alright, the homicide rates should all be descending over time. What a comfort. But the viz is not super clear. Let's check the rates for individual countries.

## Facet Homicide Trends by Country

You can visualize each country's trend separately by adding an extra argument to the ggplot, the `facet_wrap()` and feeding it the country column. If in doubt, check your ggplot tutorial and your country column name for exact usage.

- reuse the ggplot from the chunk above
- insert `facet_wrap()` after the specification of geometry to split countries in separate charts
- change the facet "layout" to two columns and three rows so that the trends are easier to see in horizontal layout. 


```{r eval = FALSE}
ggplot(Western_Europe, aes(x = Year, y = homicides_per_100k, color = Entity)) +
  geom_line() +
  facet_wrap(~Entity, ncol = 2, nrow = 3) +
  labs(
    title = "Homicide Rates by Country in Western Europe",
    x = "Year",
    y = "Homicides per 100,000 People"
  ) +
  theme_bw()
```

## Compare Homicides with Danish Monarchs’ Reigns

- Load your Danish king dataset. Hopefully it is tidy and your years and duration of reign are all numeric.
- You need to have a consistent way of plotting the rulers' reign on the x axis, so I recommend you create a midyear column by calculating the middle of each monarch's rule (Hint: `midyear` = `endyear` - (`endyear`-`startyear`)/2)
- Start a ggplot plotting midyear on x axis and duration (Periode) on y axis
- Try `geom_smooth()` for geometry 
- Provide meaningful labels and a title
- How would you characterize the trend compared to the homicides below?
Den nedadgående graf, indikerer at samfundet har ændret sig i en mere sikker retning, hvor folk i lavere grad bliver udsat for mord.

```{r eval = FALSE}
# Step 1: Load the danish monarchs dataset
Danish_Rulers <- read_csv2("data/danish_monarchs_clean.csv")

# Step 2: Inspect the date
glimpse(Danish_Rulers)

# Step 3: Create midyear column
Danish_Rulers <- Danish_Rulers %>%
  mutate(midyear = Start_regeringsaar + (Slut_regeringsaar - Start_regeringsaar) / 2)

# Step 4: Inspect the updated data
glimpse(Danish_Rulers)

# Step 5: Visualize duration of reign over time
ggplot(Danish_Rulers, aes(x = midyear, y = Periode)) +
  geom_smooth(se = FALSE, color = "blue") +
  labs(
    title = "Duration of Danish Rulers' Reigns Over Time",
    x = "Midyear of Reign",
    y = "Duration of Reign (Years)"
  ) +
  theme_bw()
```


##Final tasks:

1) Plot: In the facetted plot above, move the legend from the current position on the side to below the facets, and label it "Country" instead of "Entity".

```{r}
ggplot(Western_Europe, aes(x = Year, y = homicides_per_100k, color = "Country")) +
  geom_line() +
  facet_wrap(~Entity, ncol = 2, nrow = 3) +
  labs(
    title = "Homicide Rates by Country in Western Europe",
    x = "Year",
    y = "Homicides per 100,000 People"
  ) +
  theme_bw()+
  theme(legend.position = "bottom")
```


2) Rmarkdown: 

  - edit the author of the document, and convert 'Final Tasks' into heading #2 (like the other headings) 
  - add a `floating table of contents` to your Rmarkdown document, 
  - provide informative `chunk-names` and edit flags in your R chunks, and
  - automatically generate a `timestamp` to show when the document was last updated. (Hint: check the Rmarkdown episode in our Data Carpentry tutorial)
  
```{r timestamp, results='asis', echo=FALSE}

cat(paste("**Document last updated on:**", format(Sys.time(), "%Y-%m-%d %H:%M:%S")))

```
  

###  Final question: 

*In <250 words articulate your answer on the basis of the data visualisations to the following question*:** are we more civilized today?** *(Can you find an article or a book that discusses this topic?)*

Som udgangspunkt, hvis man kigger på grafen, er der gennem tiden færre mennesker der bliver dræbt. I 1400-tallet kan man se at i Italien bliver der dræbt over 60 mennesker per 100.000, hvor grafen det meste af tiden, udover et hop, er nedadgående ligesom resten af Europa, hvor den ligger på 1-3 mennesker pr. 100.000. Trenden tyder på at vi lever i et mere fredeligt og lovlydigt samfund, hvor vi ikke dræber hinanden, og dermed er blevet mere “civiliseret”.